# Example using pseudo Facebook data

Load data (the file as a tab separator)
```{r}
if (basename(getwd()) != 'dataset') {
  setwd('./dataset')
}
list.files()
pf = read.csv("pseudo_facebook.tsv", sep="\t")
```

See dataframe structure
```{r}
str(pf)
names(pf)
```

First visualization
```{r}
library(ggplot2)
qplot(x=dob_day, data=pf)
```

Change the x ticks
```{r}
qplot(x=dob_day, data=pf) +
  scale_x_continuous(breaks = 1:31)
```


Run the following code in R to get other themes.

```
install.packages('ggthemes', dependencies = TRUE)
library(ggthemes)
```

The ggthemes package was developed by Jeffery Arnold. Check out examples of the themes on the github page.

Chris is using theme_minimal()with the font size set to 24, which is why his output might look slightly different than yours. You can set the same theme in R by running the following code, or you can set the theme to a choice of your own.
theme_set(theme_minimal(24))


Instead of using the qplot() function, you can also use the ggplot() function to create the histogram:

```
ggplot(aes(x = dob_day), data = pf) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 1:31)
```

Try with a different theme. See more info [here](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html):
```{r}
#install.packages('ggthemes', dependencies = TRUE)
library(ggthemes)
theme_set(theme_economist())
qplot(x=dob_day, data=pf) +
  scale_x_continuous(breaks = 1:31)
```

Facets
```{r}
qplot(x=dob_day, data=pf) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month, ncol=3)
```

We can see that most people are born the first of January (this is the default birth date when creating an account on Facebook).

Let's now plot friends !
```{r}
qplot(x=friend_count, data=pf)
```

We can change the X axis using two different ways:

```{r}
qplot(x=friend_count, data=pf, xlim=c(0, 1000))
```

Or using a layer:

```{r}
qplot(x=friend_count, data=pf) +
  scale_x_continuous(limits=c(0, 1000))
```

We can also change the bin width:

```{r}
qplot(x=friend_count, data=pf, binwidth=25) +
  scale_x_continuous(limits=c(0, 1000))
```

And the x ticks (breaks)

```{r}
qplot(x=friend_count, data=pf, binwidth=25) +
  scale_x_continuous(limits=c(0, 1000), breaks=seq(0, 1000, 50))
```

We can use facet to make distinction between males and females:

```{r}
qplot(x=friend_count, data=pf, binwidth=25) +
  scale_x_continuous(limits=c(0, 1000), breaks=seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 1)
```

All of this may be written with ggplot syntax:

```{r}
ggplot(aes(x = friend_count), data = pf) +
  geom_histogram(binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 1)
```

We have one histogram with NA, we can subset the dataframe to avoid this:

```{r}
qplot(x=friend_count, data=subset(pf, !is.na(gender)), binwidth=25) +
  scale_x_continuous(limits=c(0, 1000), breaks=seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 1)
```

What is the number of friends per gender ?

```{r}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

From the histograms, we may have the feeling that males have more friends. But actually it's not the case.

We can have a look to tenure parameter with custom colors:

```{r}
qplot(x=tenure, data=pf, color=I("black"), fill=I("#099DD9"))
```

Let's tume the histogram so that it is years instead if days in x axis:

```{r}
qplot(x=tenure/365.25, data=pf, binwidth=0.25, color=I("black"), fill=I("#099DD9")) + 
  labs(x="years") +
  scale_x_continuous(breaks=c(0:8))
```

We can also create grid of plots:

```{r}
#install.packages("gridExtra")
library(gridExtra)
p1 = qplot(x=friend_count, data=pf, bins=50, color=I("black"), fill=I("#099DD9")) + 
  scale_x_continuous(limits=c(0, 1500), breaks=(seq(0, 1500, 100))) +
  labs(x="Number of friends (continuous)")
p2 = qplot(x=friend_count, data=pf, bins=50, color=I("black"), fill=I("#099DD9")) + 
  scale_x_log10(limits=c(1, 1500), breaks=c(1, 10, 50, 100, 500, 1000)) +
  labs(x="Number of friends (log10)")
p3 = qplot(x=friend_count, data=pf, bins=50, color=I("black"), fill=I("#099DD9")) +
  scale_x_sqrt(limits=c(0, 1500)) + 
  labs(x="Number of friends (sqrt)")

# arrange plots in grid
grid.arrange(p1, p2, p3, ncol=1)
```

We can use frequency plot instead of histograms:

```{r}
qplot(x=friend_count, y=..count../sum(..count..),
      data=subset(pf, !is.na(gender)), binwidth=10, 
      geom="freqpoly", color=gender) +
  scale_x_continuous(limits=c(0, 1000), breaks=seq(0, 1000, 50)) + 
  labs(y="Proportion of users with that friend count", x="Friend count")
```

We can use frequency plot for likes as well:

```{r}
qplot(x=www_likes, data=subset(pf, !is.na(gender)), geom="freqpoly", color=gender, bins=40) + 
  scale_x_log10(limits=c(1, 5000), breaks=c(1, 5, 10, 50, 100, 150, 1000, 5000))
```

Use calculus to see www_likes:

```{r}
by(pf$www_likes, pf$gender, sum)
```

Now let's use box plots:

```{r}
qplot(y=friend_count, x=gender,
      data=subset(pf, !is.na(gender)),
      geom="boxplot") + 
  scale_y_continuous(limits=c(0, 1000), breaks=seq(0, 1000, 100))
```

Problem with this code is that samples are removed with the scale_y_continuous/limits. There is an alternative:

```{r}
qplot(y=friend_count, x=gender,
      data=subset(pf, !is.na(gender)),
      geom="boxplot") + 
  coord_cartesian(ylim=c(0, 1000)) +
  scale_y_continuous(breaks=seq(0, 1000, 100))
```

Getting logical

```{r}
summary(pf$mobile_likes)
summary(pf$mobile_likes > 0)
```

Or creating a new variable:

```{r}
pf$mobile_check_in <- NA
pf$mobile_check_in <- factor(ifelse(pf$mobile_likes > 0, 1, 0))
```




