Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
library(ggplot2)
pf <- read.csv("../L03-ExploreOneVariable/dataset/pseudo_facebook.tsv", sep = "\t")
ggplot(aes(x=gender, y=age), data=subset(pf, !is.na(gender))) + 
  geom_boxplot() +
  stat_summary(fun.y = mean, geom="point", shape=3)
```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}
library(dplyr)
pf.fc_by_age_gender <- pf %>%
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>%
  summarize(mean_friend_count=mean(friend_count),
            median_friend_count=median(friend_count),
            n = n()) %>%
  arrange(age)
head(pf.fc_by_age_gender, 10)
```

```{r}
ggplot(aes(x = age),
       data = pf.fc_by_age_gender) +
  geom_line(aes(y=median_friend_count, color = gender))
```


***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

Long format > dcast > Wide format > melt > Long format

See tutorial [here](http://seananderson.ca/2013/10/19/reshape.html).

```{r}
#install.packages('reshape2')
library(reshape2)
pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender, 
                                  age ~ gender,
                                  value.var = "median_friend_count")
head(pf.fc_by_age_gender.wide)
```

```{r}
pf.fc_by_age_gender.long <- melt(pf.fc_by_age_gender.wide, 
                                 c("age"),
                                 variable.name="gender",
                                 value.name="median_friend_count")
pf.fc_by_age_gender.long <- arrange(pf.fc_by_age_gender.long, age)
head(pf.fc_by_age_gender.long)
```


***

### Ratio Plot
Notes:

```{r Ratio Plot}
pf.fc_by_age_gender.wide$ratio <- with(pf.fc_by_age_gender.wide, female / male)
ggplot(aes(x = age),
       data = pf.fc_by_age_gender.wide) +
  geom_line(aes(y=ratio)) +
  geom_hline(yintercept=1.0, linetype=2)
```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
pf$year_joined <- with(pf, floor(2014 - tenure / 365))
```

***

### Cut a Variable
Notes:

```{r Cut a Variable}
summary(pf$year_joined)
table(pf$year_joined)
pf$year_joined.bucket <- cut(pf$year_joined, breaks=c(2004, 2009, 2011, 2012, 2014))
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
table(pf$year_joined.bucket, useNA = "ifany")
ggplot(aes(x=age, y=friend_count), 
       data=subset(pf, !is.na(pf$year_joined.bucket))) +
  geom_line(aes(color=year_joined.bucket), stat="summary", fun.y=median)
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}
ggplot(aes(x=age, y=friend_count), 
       data=subset(pf, !is.na(pf$year_joined.bucket))) +
  geom_line(aes(color=year_joined.bucket), stat="summary", fun.y=mean) +
  geom_line(stat="summary", fun.y=mean, linetype=2)
```

***

### Friending Rate
Notes:

```{r Friending Rate}
pf$friending_rate <- with(pf, ifelse(tenure > 0, friend_count / tenure, NA))
summary(pf$friending_rate)
```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}
pf$friendships_rate <- with(pf, ifelse(tenure > 0, friendships_initiated / tenure, NA))
ggplot(aes(x=tenure, y=friendships_rate), 
       data=subset(pf, tenure > 0)) +
  geom_line(aes(color=year_joined.bucket), stat="summary", fun.y=mean)
```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_smooth(aes(color = year_joined.bucket))
```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
yo <- read.csv("dataset/yogurt.csv")

#change id variable to a factor (categorical data)
yo$id <- factor(yo$id)
str(yo)

ggplot(aes(x=price), data=yo) +
  geom_histogram()
```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)
ggplot(aes(x=all.purchases), data=yo) +
  geom_histogram(binwidth=1.0)
```

***

### Prices over Time
Notes:

```{r Prices over Time}
ggplot(aes(x=time), data=yo) +
  geom_point(aes(y=price), alpha=0.1, position=position_jitter(width=5, height=1.))
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(4230)
samples.id <- sample(levels(yo$id), 16)
ggplot(aes(x=time, y=price), 
       data=subset(yo, id %in% samples.id)) +
  facet_wrap(~id) + 
  geom_line() + 
  geom_point(aes(size=all.purchases), pch=1)
```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("dataset/nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

