# Problem set: explore many variables

## Load data

```{r}
library(ggplot2) #must load the ggplot package first
data(diamonds) #loads the diamonds data set since it comes with the ggplot package
summary(diamonds)
str(diamonds)
?diamonds
```

## Price histogram

```{r}
ggplot(aes(x=price), data=diamonds) +
  geom_histogram(aes(color=cut, fill=cut), position="stack") +
  scale_y_sqrt() + 
  facet_wrap(~color) 
```

## Price vs table

```{r}
ggplot(aes(x=table), data=diamonds) +
  geom_point(aes(y=price, color=cut), position=position_jitter(width=0.3), alpha=0.5) +
  scale_x_continuous(breaks=seq(52, 64, 1), limits=c(quantile(diamonds$table, 0.01), quantile(diamonds$table, 0.99)))
```

## Price vs volume

```{r}
diamonds$volume = with(diamonds, x * y *z)
ggplot(aes(x=volume), data=diamonds) +
  geom_point(aes(y=price, color=clarity), alpha=0.5) +
  scale_x_continuous(limits=c(quantile(diamonds$volume, 0.01), quantile(diamonds$volume, 0.99))) + 
  scale_y_log10()
```

## Proportion of friendships initiated

```{r}
pf <- read.csv("../L03-ExploreOneVariable/dataset/pseudo_facebook.tsv", sep = "\t")
pf$prop_initiated <- with(pf, ifelse(friend_count > 0, friendships_initiated / friend_count, NA))
head(pf, 10)
```

# Create a line graph of the median proportion of
# friendships initiated ('prop_initiated') vs.
# tenure and color the line segment by
# year_joined.bucket.

# Recall, we created year_joined.bucket in Lesson 5
# by first creating year_joined from the variable tenure.
# Then, we used the cut function on year_joined to create
# four bins or cohorts of users.

# (2004, 2009]
# (2009, 2011]
# (2011, 2012]
# (2012, 2014]

```{r}
pf$year_joined <- with(pf, floor(2014 - tenure / 365))
pf$year_joined.bucket <- cut(pf$year_joined, breaks=c(2004, 2009, 2011, 2012, 2014))
ggplot(aes(x=tenure), data=pf) +
  geom_line(aes(y=prop_initiated, color=year_joined.bucket), stat="summary", fun.y=median)
```
```{r}
pf$year_joined <- with(pf, floor(2014 - tenure / 365))
pf$year_joined.bucket <- cut(pf$year_joined, breaks=c(2004, 2009, 2011, 2012, 2014))
ggplot(aes(x=6*round(tenure/6)), data=pf) +
  geom_line(aes(y=prop_initiated, color=year_joined.bucket), stat="summary", fun.y=median)
```

```{r}
ggplot(aes(x=tenure), data=pf) +
  geom_smooth(aes(y=prop_initiated, color=year_joined.bucket))
```

```{r}
summary(subset(pf, year_joined.bucket == "(2012,2014]")$prop_initiated)
```

## Price vs carat

# Create a scatter plot of the price/carat ratio
# of diamonds. The variable x should be
# assigned to cut. The points should be colored
# by diamond color, and the plot should be
# faceted by clarity.

```{r}
ggplot(aes(x=cut), data=diamonds) +
  geom_point(aes(y=price/carat, color=color), position=position_jitter(), alpha=0.6) +
  facet_wrap(~clarity)
```


