Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
library(ggplot2)
pf <- read.csv("../L03-ExploreOneVariable/dataset/pseudo_facebook.tsv", sep = "\t")
qplot(x=age, y=friend_count, data=pf)
```

***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x=age, y=friend_count), data=pf) + 
  geom_point() + 
  xlim(13, 90)
summary(pf$age)
```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(aes(x=age, y=friend_count), data=pf) + 
  geom_jitter(alpha=1/20) + 
  xlim(13, 90)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(aes(x=age, y=friend_count), data=pf) + 
  geom_point(alpha=1/20) + 
  xlim(13, 90) +
  coord_trans(y='sqrt')
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
ggplot(aes(x=age, y=friend_count), data=pf) + 
  geom_point(alpha=1/20, position=position_jitter(height = 0)) + 
  xlim(13, 90) +
  coord_trans(y='sqrt')
```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x=age, y=friendships_initiated), data=pf) + 
  geom_jitter(alpha=1/20, position=position_jitter(height = 0)) + 
  xlim(13, 90) +
  coord_trans(y='sqrt')
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
library(dplyr)
age_groups <- group_by(pf, age)
pf.fc_by_age <- summarise(age_groups,
                          friend_count_mean = mean(friend_count),
                          friend_count_median = median(friend_count),
                          friend_count_count = n())
pf.fc_by_age <- arrange(pf.fc_by_age, age)
head(pf.fc_by_age)
```

```{r Conditional Means Alternate method}
library(dplyr)

pf.fc_by_age <- pf %>%
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            friend_count_count = n()) %>%
  arrange(age)
head(pf.fc_by_age)
```

Create your plot!

```{r Conditional Means Plot}
ggplot(aes(x=age), data=pf.fc_by_age) + 
  geom_line(aes(y=friend_count_median, color="median")) +
  geom_line(aes(y=friend_count_mean, color="mean")) +
  labs(y="Friend count")
```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x=age, y=friend_count), data=pf) + 
  geom_jitter(alpha=1/20, position=position_jitter(height = 0), color="orange") +
  geom_line(stat='summary', fun.y=mean) +
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.1), color='blue', linetype = 2) + 
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.5), color='green', linetype = 2) + 
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.9), color='red', linetype = 2) + 
  coord_cartesian(xlim=c(13, 75), ylim=c(0, 1000))
```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor.test(pf$age, pf$friend_count, method="pearson")
with(pf, cor.test(age, friend_count, method="pearson"))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(pf[pf$age <= 70,] , cor.test(age, friend_count, method="pearson"))
with(pf[pf$age <= 70,] , cor.test(age, friend_count, method="spearman"))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(aes(x=www_likes_received, y=likes_received), data=pf) + 
  geom_point(alpha=1/20) +
  coord_trans(y='sqrt', x='sqrt')
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(x=www_likes_received, y=likes_received), data=pf) + 
  geom_point(alpha=1/20) +
  #coord_trans(y='sqrt', x='sqrt') +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) + 
  geom_smooth(method = "lm")
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pf , cor.test(www_likes_received, likes_received, method="pearson"))
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
#install.packages('alr3')
library(alr3)
data("Mitchell")
?Mitchell
```

Create your plot!

```{r Temp vs Month}
ggplot(aes(x=Month, y=Temp), data=Mitchell) + 
  geom_point()
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}
with(Mitchell, cor.test(Month, Temp, method="pearson"))
```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}
ggplot(aes(x=Month, y=Temp), data=Mitchell) + 
  geom_point() +
  scale_x_continuous(breaks=seq(0, 203, 12))
```
```{r Making Sense of Data}
ggplot(aes(x=Month%%12, y=Temp), data=Mitchell) + 
  geom_point()
```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
ggplot(aes(x=age), data=pf.fc_by_age) + 
  geom_line(aes(y=friend_count_median, color="median")) +
  geom_line(aes(y=friend_count_mean, color="mean")) +
  labs(y="Friend count")
```

***

### Age with Months Means

```{r Age with Months Means}
pf$age_with_months <- with(pf, age + (1 - dob_month / 12))
```

Programming Assignment
```{r Programming Assignment}
pf.fc_by_age_with_month <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            friend_count_count = n()) %>%
  arrange(age_with_months)
head(pf.fc_by_age_with_month)
```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
ggplot(aes(x=age_with_months), data=subset(pf.fc_by_age_with_month, age_with_months < 71)) + 
  geom_line(aes(y=friend_count_median, color="median")) +
  geom_line(aes(y=friend_count_mean, color="mean")) +
  labs(y="Friend count")
```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}
p1 <- ggplot(aes(x=age_with_months), data=subset(pf.fc_by_age_with_month, age_with_months < 71)) + 
  geom_line(aes(y=friend_count_mean)) +
  geom_smooth(aes(y=friend_count_mean)) +
  labs(y="Friend count")
p2 <-ggplot(aes(x=age), data=pf.fc_by_age) + 
  geom_line(aes(y=friend_count_mean)) +
  geom_smooth(aes(y=friend_count_mean)) +
  labs(y="Friend count")
library(gridExtra)
grid.arrange(p1, p2, ncol = 1)
```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

